Paper

### 2019-12-09 Paper_resubmit.Rmd
cleaning things up

qq = quantile(features$hazards, seq(0,1,.15))
p = ggplot(features, aes(coef, hazards, col=hazards >= qq[length(qq)])) + 
  geom_point(alpha=0.8, show.legend = F) + 
  geom_text_repel(aes(x=coef, label=ifelse(hazards >= qq[length(qq)], label, '')),show.legend=F) +
  scale_color_manual(values=c('firebrick3','darkblue')) +
  labs(title='Top features by cv(RR)', x='Coefficient value', y='cvRR') + plot.theme 
p
ggsave('plots/haz_coef.png', plot=p, height=5, width=5, units='in', dpi=300)
```

@@

```{r leaveoneout, echo=F, message=F, warning=F, fig.width=5, fig.height=5}
file = paste(cache.dir, 'loo_0.9.Rdata', sep='/')
if (file.exists(file)) {
  load(file, verbose=F)
} else {
  stop(paste("Missing model file", file))
}
loo.coefs = nzcoefs

@@

```{r prediction_cutoff, echo=F, message=F, warning=F, fig.height=7, fig.width=7}
preds = pg.samp %>% dplyr::select(Status, Prediction)
roc = pROC::roc(Status ~ Prediction, data=preds, auc=T, ci=T, of='thresholds',transpose=T)
ggsave(filename='plots/auc/training_auc.png', plot=roc.plot(roc), width=6.5, height=4.5, units='in', dpi=300)

@@

@@

pathsamples = patient.info[,c('Pathology','Status')]
pathroc = pROC::roc((as.integer(pathsamples$Status)-1), ifelse(pathsamples$Pathology %in% c('LGD','HGD','IMC'), 1, 0) )
pathroc$model = 'HGD & LGD'
ggsave('plots/auc/dysplasia_auc.png', plot=roc.plot(pathroc, title='Dysplasia (HGD & LGD)'), width=6, height=6, units="in", limitsize=F, dpi = 300)

@@

```{r, echo=F, warning=F, message=F, fig.height=10, fig.width=12}
p53Path = pg.samp %>% dplyr::filter(Status == 'P' & !is.na(p53.Status)) %>% 
  dplyr::group_by(Pathology,p53.Status) %>% 
  dplyr::summarise(Freq = length(Pathology)) %>% dplyr::mutate(Status = 'p53 IHC')

@@

rp = plot.risk.by.path(pg.samp)
rp
ggsave(filename='plots/pred_by_path.png', plot=(rp + theme(legend.position = 'none')), width=7, height=6, units='in', dpi=300)
```

@@

Using the discretized risk what is the AUC?

@@

@@

@@

@@

```{r}
roc = pROC::roc(Status ~ Probability, data=val.preds, auc=T, ci=T, of='thresholds',transpose=T)
roc.plot(roc) + labs(title='Validation cohort')
ggsave('plots/validation/validatation_roc.png', plot=roc.plot(roc) + labs(title='Validation cohort'), width=5, height=5, units='in',dpi=300)

@@

```{r}
rp = plot.risk.by.path(val.preds) + labs(title='Validation samples predicted by pathology')  
rp
ggsave(filename='plots/validation/val_pred_by_path.png', plot=rp, width=5, height=6, units='in', dpi=300)

@@

kmp = survminer::ggsurvplot(sfit, size=1, linetype="strata",risk.table=T, risk.table.col='strata',
                      conf.int=T, censor=T, legend='none', title='Progression risk over time', xlab='Time (months prior to progression)',
                      ylab = 'Progression risk',
                      ggtheme=theme_minimal(), palette=as.vector(BarrettsProgressionRisk::riskColors()), font.legend=c(14,'bold'))

@@

kmp = survminer::ggsurvplot(sfit, size=1, linetype="strata",risk.table=T, risk.table.col='strata',
                      conf.int=T, censor=T, legend='none', title='Progression Risk', 
                      subtitle='Training & Validation predictions', xlab='Time (months prior to progression)',
                      ylab = 'Progression risk',
                      ggtheme=theme_minimal(), palette=as.vector(BarrettsProgressionRisk::riskColors()), font.legend=c(14,'bold'))

@@

ggsave('plots/risk_km.png', plot=print(kmp), height=9, width=8, units = 'in', dpi = 300)

@@

```{r, rx, echo=F, warning=F, message=T, fig.width=8, fig.height=5}

rx<-function(Patient, brks, Prediction, Pathology, Risk, p53.Status) {
  rules = as.data.frame(matrix(ncol=0, nrow=0))
  # Consecutive
  for (i in 1:length(brks)) {
    risks = table(Risk[i:(i+1)])
    p53 = table(p53.Status[i:(i+1)])
    rule = 'None'
    if ( risks['High'] == 2 | length(which(grepl('HGD|IMC', Pathology[i:(i+1)]))) > 0 ) {
      rule = 1
    } else if ( risks['High'] == 1 | p53['1'] > 0 ) {
      rule = 2
    } else if ( risks['Moderate'] > 0 ) {
      rule = 3
    } else if ( risks['Low'] == 2) {
      rule = 4
    }
  rules = rbind(rules, cbind( as.character(brks[i]), as.character(brks[(i+1)]), rule))
  }  
  colnames(rules) = c('t1','t2', 'rule')
  rules$Patient = Patient
  rules
}

pg.samp = pg.samp %>% mutate(p53.Status = factor(p53.Status, levels=c(0,1), ordered=T))

maxP = pg.samp %>% dplyr::group_by(Patient, Status, brks) %>% 
  dplyr::summarise( 'maxP'=max(Prediction), 'Risk'=max(Risk), 'Pathology'=max(Pathology), 'p53'=max(p53.Status,na.rm=T) )

pt.counts = maxP %>% group_by(Patient, Status) %>% dplyr::filter( length(which(grepl('Endpoint', brks))) > 0  & length(brks) > 1) %>% 
  ungroup %>% dplyr::select(Patient, Status) %>% distinct %>% group_by(Status) %>% tally

rules = do.call(bind_rows, purrr::map(unique(maxP$Patient), function(pt) {
  r = with(arrange(subset(maxP, Patient == pt), brks), rx(pt, brks,maxP,Pathology,Risk,p53))
  r$Status = unique(subset(maxP, Patient == pt)$Status)
  return(as_tibble(r))
})) %>% dplyr::mutate(t1 = factor(t1, levels=levels(pg.samp$brks), ordered=T), t2 = factor(t2, levels=levels(pg.samp$brks), ordered=T)) %>% arrange(Patient, t2)

rules %>% group_by(Patient, Status) %>% dplyr::filter(t2 == 'Endpoint' & length(t1) > 2) %>% 
  dplyr::select(Patient, Status) %>% distinct %>%  group_by(Status) %>% tally

ids = rules %>% group_by(Patient, Status) %>% dplyr::filter(t2 == 'Endpoint' & length(t1) > 2) %>% dplyr::select(Patient) %>% pull

penult = rules %>% dplyr::filter(Patient %in% ids & t2 != 'Endpoint') %>% arrange(Patient, desc(t1)) %>% group_by(Patient, Status) %>% 
  dplyr::mutate(gid = 1:n() ) %>% dplyr::filter(gid == 1)

#penult = filter(rules, t2 != 'Endpoint' & (t1 %in% rev(levels(pg.samp$brks))[2] | t2 == rev(levels(pg.samp$brks))[2] ))

penult %>% dplyr::select(Patient, Status) %>% distinct %>% group_by(Status) %>% tally

penult %>% filter(Status == 'P' & rule %in% c(3,4)) %>% kable(caption='Patients that would be missed if we moved Moderate to Low risk') %>% kable_styling(full_width = F)

penult %>% filter(Status == 'NP' & rule == 1) %>% kable(caption='NP patients for RFA') %>% kable_styling(full_width = F)

penult %>% group_by(Status) %>% tally(name='total')

penult %>% group_by(Status, rule) %>% tally %>% spread(rule,n) %>% 
  kable(caption='Penult Endoscopy: # patients by rx rules based on the max (P) per endoscopy') %>% kable_styling(full_width=F)

penult = rules %>% dplyr::filter(Patient %in% ids & t2 != 'Endpoint') %>% arrange(Patient, desc(t1)) %>% group_by(Patient, Status) %>% 
  dplyr::mutate(gid = 1:n() ) %>% dplyr::filter(gid == 1)

path.rx<-function(Patient, brks, Pathology) {
  rules = as.data.frame(matrix(ncol=0, nrow=0))
  # Consecutive
  for (i in 1:length(brks)) {
    #risks = table(Pathology[i:(i+1)])
    rule = 'None'
    if ( length(which(grepl('HGD|IMC', Pathology[i:(i+1)]))) > 0  |
         length(which(grepl('LGD', Pathology[i:(i+1)]))) > 1 ) {
      rule = 1
    } else if ( length(which(grepl('LGD', Pathology[i:(i+1)]))) > 0 ) {
      rule = 2
    } else if ( length(which(grepl('BE|NDBE|ID', Pathology[i:(i+1)]))) > 1  ) {
      rule = 3
    }
  rules = rbind(rules, cbind( as.character(brks[i]), as.character(brks[(i+1)]), rule))
  }  
  colnames(rules) = c('t1','t2', 'rule')
  rules$Patient = Patient
  rules
}

rules2 = do.call(rbind, lapply(unique(maxP$Patient), function(pt) {
  r = with(arrange(subset(maxP, Patient == pt), brks), path.rx(pt, brks,Pathology))
  r$Status = unique(subset(maxP, Patient == pt)$Status)
  r
}))
rules2$t1 = factor(rules2$t1, levels=levels(pg.samp$brks), ordered=T)
rules2$t2 = factor(rules2$t2, levels=levels(pg.samp$brks), ordered=T)

second = subset(rules2,  t2 != 'Endpoint') %>% 
  dplyr::group_by(Patient, Status) %>% 
  dplyr::summarise(  max(t2), 'rule'=rule[t2 == max(t2)]  )

second %>% group_by(Status, rule) %>% tally %>% spread(rule,n) %>% 
  kable(caption='Patient recommendation by PATH alone in penult endoscopy' ) %>% kable_styling(full_width = F)

second %>% group_by(Status, rule) %>% tally %>% spread(rule,n) %>% 
  left_join(second %>% group_by(Status) %>% tally(name='total'), by='Status') %>%
  mutate_if(is.numeric, funs(  round((./total)*100,1)  )) %>% dplyr::select(-total) %>%
  kable(caption='Final endoscopy: % patients by recommendation rule based on the max (PATH) per endoscopy, not (RISK)') %>% kable_styling(full_width=F)

@@

>>>>

qq = quantile(features$hazards, seq(0,1,.15))
p = ggplot(features, aes(coef, hazards, col=hazards >= qq[length(qq)])) + 
  geom_point(alpha=0.8, show.legend = F) + 
  geom_text_repel(aes(x=coef, label=ifelse(hazards >= qq[length(qq)], label, '')),show.legend=F) +
  scale_color_manual(values=c('firebrick3','darkblue')) +
  labs(title='Top features by cv(RR)', x='Coefficient value', y='cvRR') + plot.theme 
p
ggsave('plots/haz_coef.png', plot=p, height=5, width=5, units='in', dpi=300)

cvRR %>% mutate(`Gain/Loss` = ifelse(coef > 0 ,'Gain', 'Loss')) %>% 
  dplyr::select(label, hazards, `Gain/Loss`) %>% arrange(desc(hazards)) %>%
  write_tsv('plots/coef_effect.tsv')

@@

```{r leaveoneout, echo=F, message=F, warning=F, fig.width=5, fig.height=5}
# loo_residuals = do.call(bind_rows, purrr::map(list.files(paste0(dir, '/Analysis/pcf_perPatient/50kb'), 'residuals.tsv', recursive = T, full.names = T), function(f) {
#   read_tsv(f, col_types = 'ccddddl')
# }))
# 
# failedQC = loo_residuals %>% filter(varMAD_median > 0.008)

file = paste(cache.dir, 'loo_0.9.Rdata', sep='/')
if (file.exists(file)) {
  load(file, verbose=F)
} else {
  stop(paste("Missing model file", file))
}
loo.coefs = nzcoefs

@@

```{r prediction_cutoff, echo=F, message=F, warning=F, fig.height=7, fig.width=7}
preds = pg.samp %>% dplyr::select(Status, Prediction)
roc = pROC::roc(Status ~ Prediction, data=preds, auc=T, ci=T, of='thresholds',transpose=T)

pROC::ci.auc(roc$auc)

ggsave(filename='plots/auc/training_auc.png', plot=roc.plot(roc), width=6.5, height=4.5, units='in', dpi=300)

@@

pROC::ci.auc(p53roc$auc)

@@

pathsamples = patient.info[,c('Pathology','Status')]
pathroc = pROC::roc((as.integer(pathsamples$Status)-1), ifelse(pathsamples$Pathology %in% c('LGD','HGD','IMC'), 1, 0) )
pROC::ci.auc(pathroc$auc)

pathroc$model = 'HGD & LGD'
ggsave('plots/auc/dysplasia_auc.png', plot=roc.plot(pathroc, title='Dysplasia (HGD & LGD)'), width=6, height=6, units="in", limitsize=F, dpi = 300)

@@

#### By pathology?
```{r, echo=F, warning=F, message=F, fig.height=10, fig.width=12}
p53Path = pg.samp %>% dplyr::filter(Status == 'P' & !is.na(p53.Status)) %>% 
  dplyr::group_by(Pathology,p53.Status) %>% 
  dplyr::summarise(Freq = length(Pathology)) %>% dplyr::mutate(Status = 'p53 IHC')

@@

rp = plot.risk.by.path(pg.samp %>% ungroup)
rp
ggsave(filename='plots/pred_by_path.png', plot=(rp + theme(legend.position = 'none')), width=7, height=6, units='in', dpi=300)
```

@@

#### Using the discretized risk what is the AUC?

@@

One way of looking at the p53 prediction (e.g. Adam Bass) is to see how predictive the p-arm loss in chr17 is.

@@

pi %>%  mutate(p53.Status = recode_factor(p53.Status, '1'='aberrant','0'='normal')) %>% dplyr::group_by(p11Loss, p53.Status) %>% tally %>% spread(p53.Status, n)

@@

## q arm
```{r, echo=T}

q11 = filter(globalCNdf, chr == 17 & start > 1 & seg.type == 'arms')
q11gain = q11 %>% filter(value > 1)

q11gain %>% filter(value >= 1) %>% group_by(Status) %>% tally(name='q11 Gain') %>% 
  full_join( q11 %>% group_by(Status) %>% tally(name='Total Samples') ) %>% 
  kable(caption='Chr17 q11 gain') %>% kable_styling(full_width = F)

q11gain %>% filter(value >= 1) %>% dplyr::select(Patient, Status) %>% distinct %>%
  group_by(Status) %>%  tally(name='q11 Gain') %>%
  kable(caption='Patients showing q gain') %>% kable_styling(full_width = F)

q11Path = patient.info %>% filter(Samplename %in% q11gain$variable) %>% dplyr::group_by(Pathology) %>% tally(name='samples')

q11Path %>% dplyr::mutate('ratio'=round(samples/sum(q11Path$samples),2)) %>% 
kable(caption='Ratio of all Chr17 q11 gain samples by pathology' ) %>% kable_styling(full_width = F)
```

@@

```{r}
roc = pROC::roc(Status ~ Probability, data=val.preds, auc=T, ci=T, of='thresholds',transpose=T)
roc.plot(roc) + labs(title='Validation cohort')
pROC::ci.auc(roc$auc)

ggsave('plots/validation/validatation_roc.png', plot=roc.plot(roc) + labs(title='Validation cohort'), width=5, height=5, units='in',dpi=300)

@@

```{r}
rp = plot.risk.by.path(val.preds) + labs(title='Validation samples classified by pathology')  
rp
ggsave(filename='plots/validation/val_pred_by_path.png', plot=rp, width=5, height=6, units='in', dpi=300)

@@

kmp = survminer::ggsurvplot(sfit, size=1, linetype="strata",risk.table=T, risk.table.col='strata',
                      conf.int=T, censor=T, legend='none', title='Risks over time', xlab='Time (months prior to final sample)',
                      ylab = 'Progression risk',
                      ggtheme=theme_minimal(), palette=as.vector(BarrettsProgressionRisk::riskColors()), font.legend=c(14,'bold'))

@@

kmp = survminer::ggsurvplot(sfit, size=1, linetype="strata",risk.table=T, risk.table.col='strata',
                      conf.int=T, censor=T, legend='none', 
                      title='Risks over time', 
                      subtitle='Training & Validation predictions',
                      xlab='Time (months prior to final sample)',
                      ylab = 'Progression risk',
                      ggtheme=theme_minimal(), palette=as.vector(BarrettsProgressionRisk::riskColors()), font.legend=c(14,'bold'))

@@

ggsave('plots/risk_km.png', plot=print(kmp), height=8, width=7, units = 'in', dpi = 300)

@@

#### From the penultimate endoscopy
```{r, rx, echo=F, warning=F, message=T, fig.width=8, fig.height=5}
rx<-function(Patient, brks, Prediction, Pathology, Risk, p53.Status) {
  rules = as.data.frame(matrix(ncol=0, nrow=0))
  # Consecutive
  for (i in 1:length(brks)) {
    risks = table(Risk[i:(i+1)])
    p53 = table(p53.Status[i:(i+1)])
    rule = 'None'
    if ( risks['High'] == 2 | length(which(grepl('HGD|IMC', Pathology[i:(i+1)]))) > 0 ) {
      rule = 1
    } else if ( risks['High'] == 1 | p53['1'] > 0 ) {
      rule = 2
    } else if ( risks['Moderate'] > 0 ) {
      rule = 3
    } else if ( risks['Low'] == 2) {
      rule = 4
    }
  rules = rbind(rules, cbind( as.character(brks[i]), as.character(brks[(i+1)]), rule))
  }  
  colnames(rules) = c('t1','t2', 'rule')
  rules$Patient = Patient
  rules
}

pg.samp = pg.samp %>% mutate(p53.Status = factor(p53.Status, levels=c(0,1), ordered=T), Probability = Prediction)

max.rules = purrr::map(unique(pg.samp$Patient), function(pt) {
    preds = pg.samp %>% filter(Patient == pt) %>% arrange(Endoscopy.Year, months.before.final)
    endo = BarrettsProgressionRisk:::.setUpRxTablePerEndo(preds %>% dplyr::mutate(Endoscopy = months.before.final, Sample = 'Samplename'), 'max', verbose=F) %>% mutate(Endoscopy = Endoscopy*-1)
    rx = BarrettsProgressionRisk:::.apply.rules(endo, 'Endoscopy', pred.confidence )

    left_join(rx, endo %>% dplyr::select(Endoscopy, months.before.final, Endoscopy.Year, Probability), by=c('Time 1'='Endoscopy')) %>%
      mutate(`Time 2` = ifelse(months.before.final <= 0,'Endpoint',`Time 2`)) %>% 
      mutate(Patient = pt) %>% mutate_at( vars(matches('Rule|Time')), funs(as.character) ) %>%
      dplyr::select(Patient, months.before.final, Endoscopy.Year, Probability, everything()) 
})
names(max.rules) = unique(pg.samp$Patient)

progressors = max.rules[as.character(sum.patient.data %>% filter(Status == 'P') %>% dplyr::select(Patient) %>% pull)]
nonprog = max.rules[as.character(sum.patient.data %>% filter(Status == 'NP') %>% dplyr::select(Patient) %>% pull)]

pids = names(which(sapply(progressors, function(x) {
  nrow(x) >= 1 & grepl('Endpoint', x[nrow(x),'Time 2'])
})))

nids = names(which(sapply(nonprog, function(x) {
  nrow(x) >= 1 & grepl('Endpoint', x[nrow(x),'Time 2'])
})))

penult = bind_rows( do.call(bind_rows, progressors[pids]) %>% mutate(Status = 'P'),
                    do.call(bind_rows, nonprog[nids]) %>% mutate(Status = 'NP') )

penult = penult %>% filter(`Time 2` != 'Endpoint') %>% arrange(Patient, months.before.final) %>% group_by(Patient, Status) %>% dplyr::mutate(gid = 1:n() ) %>% dplyr::filter(gid == 1)

penult %>% dplyr::select(Patient, Status) %>% distinct %>% group_by(Status) %>% tally %>% kable(caption='n patients with at least two consecutive endoscopies prior to endpoints') %>%
  kable_styling(full_width = F)

penult %>% filter(Status == 'P' & Rule %in% c(3,4)) %>% kable(caption='Prog patients that would be missed if we moved Moderate to Low risk') %>% kable_styling(full_width = F)

penult %>% filter(Status == 'NP' & Rule == 1) %>% kable(caption='NP patients for RFA') %>% kable_styling(full_width = F)

penult %>% group_by(Status, Rule) %>% tally %>% spread(Rule,n) %>% 
  kable(caption='Penult Endoscopy: # patients by rx rules based on the max (P) per endoscopy') %>% kable_styling(full_width=F)

@@

#### From the second endoscopy

```{r}
pids = names(which(sapply(progressors, function(x) {
  nrow(x) > 1 & !grepl('Endpoint', x[2,'Time 2']) & !grepl('None', x[2,'Rule'])
})))

nids = names(which(sapply(nonprog, function(x) {
  nrow(x) > 1 & !grepl('Endpoint', x[2,'Time 2']) & !grepl('None', x[2,'Rule'])
})))

second = bind_rows( do.call(bind_rows, progressors[pids]) %>% mutate(Status = 'P'),
                    do.call(bind_rows, nonprog[nids]) %>% mutate(Status = 'NP') )

second = second %>% group_by(Patient, Status) %>% arrange(`Time 2`) %>% 
  dplyr::mutate(gid = 1:n() ) %>% dplyr::filter(gid == 2)

second %>% dplyr::select(Patient, Status) %>% distinct %>% group_by(Status) %>% tally %>% kable(caption='n patients with at least two years follow-up') %>%
  kable_styling(full_width = F)

second %>% group_by(Status, Rule) %>% tally %>% spread(Rule,n) %>% 
  kable(caption='Second Endoscopy: # patients by rx rules based on the max (P) per endoscopy') %>% kable_styling(full_width=F)

second %>% filter(Status == 'P' & Rule == '1') %>% ungroup %>% filter(months.before.final < 120) %>%
  dplyr::summarise_at(vars(months.before.final), list(~min(.), ~max(.), ~mean(.), ~median(.)))

second %>% filter(Status == 'NP' & Rule == '1') %>% ungroup  %>%
  dplyr::summarise_at(vars(months.before.final), list(~min(.), ~max(.), ~mean(.), ~median(.)))
